Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-19 latex/main.tex
Updates from Overleaf

Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM 
(~5 CPU years; Figure~\ref{fig:inference_resources}).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 356 have a minimum of four loci supporting the non-focal parent on both sides of the breakpoint
(we refer to these as ``robust'' recombination events,
Figure~\ref{fig:recombinant_qc}; see STAR Methods)
reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

@@

Distinguishing a truly emerging outbreak from different forms of
correlated error is non-trivial, and sc2ts has 5 parameters to fine
tune the behaviour. Firstly, we require at least 10 samples in the
group to ensure there is sufficient support over a time window of 
7 days. Then, we build a local tree of the samples and consider
some tree metrics to evaluate whether the samples within the 
group are closely related. If we have at least 2 mutations 
shared by all samples, there's at most 2 recurrent mutations
within the group and at most 5 mutations unique to each sample, 
the retrospective group is accepted for inclusion.
Although this process is highly hueristic, it successfully 
captured the majority of saltational lineages
such as Alpha (Figure~\ref{fig:subgraph_alpha}), and the 
vast majority of samples were added following the standard
daily sample grouping mechanisms. 
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_pr_pp_mp_aph_bps_pango_dated_recombinants_report.ipynb
In total, 882,398 daily 
sample groups were added to the ARG, while only 100 were
added via this retrospective grouping mechanism.

@@

\begin{figure}
\centering
% NOTE the size of the second "m" column doesn't matter, but we have to fix the
% size in order for vertical centering to work :sigh:
% The layout here isn't ideal, but it's incredibly fiddly to get center
% alignment veritically with cells to work, so this is good enough.
\begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
Q1 & \includegraphics[height=\cprheight]{static/RE_node-QCpass-427863.png}\\
Q2 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-748991.png}\\
Q3 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-411345.png}\\
Q4 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-663484.png}\\
\end{tabularx}
\caption{
Illustrative copying patterns.
Q1 to Q4 are examples drawn from the labeled quadrants in Figure~\ref{fig:recombinant_qc}.
Q1 depicts the copying pattern of a recombination event that marginally passed QC,
whereas the others depict the copying patterns of recombination events that failed QC.
}
\label{fig:copying_pattern_examples}
\end{figure}

>>>>

Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM 
(~5 CPU years; Figure~\ref{fig:inference_resources}).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 356 have at least four loci supporting the non-focal parent on both sides of the breakpoint
(we refer to these as ``robust'' recombination events,
Figure~\ref{fig:recombinant_qc}; see STAR Methods)
reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

@@

Distinguishing a truly emerging outbreak from different forms of
correlated error is non-trivial, and sc2ts has 5 parameters to fine
tune the behaviour. Firstly, we require at least 10 samples in the
group to ensure there is sufficient support over a time window of 
7 days. Then, we build a local tree of the samples and consider
some tree metrics to evaluate whether the samples within the 
group are closely related. If we have at least 2 mutations 
shared by all samples, there's at most 2 recurrent mutations
within the group and at most 5 mutations unique to each sample, 
the retrospective group is accepted for inclusion.
Although this process is highly heuristic, it successfully 
captured the majority of saltational lineages
such as Alpha (Figure~\ref{fig:subgraph_alpha}), and the 
vast majority of samples were added following the standard
daily sample grouping mechanisms. 
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_pr_pp_mp_aph_bps_pango_dated_recombinants_report.ipynb
In total, 882,398 daily 
sample groups were added to the ARG, while only 100 were
added via this retrospective grouping mechanism.

@@

\begin{figure}
\centering
% NOTE the size of the second "m" column doesn't matter, but we have to fix the
% size in order for vertical centering to work :sigh:
% The layout here isn't ideal, but it's incredibly fiddly to get center
% alignment veritically with cells to work, so this is good enough.
\begin{tabularx}{0.85\textwidth}{>{\centering\arraybackslash}m{1cm}>{\arraybackslash}m{1cm}}
Q1 & \includegraphics[height=\cprheight]{static/RE_node-QCpass-427863.png}\\
Q2 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-748991.png}\\
Q3 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-411345.png}\\
Q4 & \includegraphics[height=\cprheight]{static/RE_node-QCfail-663484.png}\\
\end{tabularx}
\caption{
Illustrative copying patterns.
Q1 to Q4 are examples drawn from the labeled quadrants in
Figure~\ref{fig:recombinant_qc}.
Q1 depicts the copying pattern of a recombination event that marginally passed QC,
whereas the others depict the copying patterns of recombination events that failed QC.
The node ID of a recombination event is shown in the middle row.
}
\label{fig:copying_pattern_examples}
\end{figure}
